---
title: "Arthritis Clinical Trial"
author: "Peiyu Li"
date: "2022-11-26"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Description
Data are from a clinical trial comparing auranofin therapy (3 mg of oral gold, twice daily) and placebo for the treatment of rheumatoid arthritis (Bombardier et al., 1986). In this six-month, randomized, double-blind trial, 303 patients with classic or definite rheumatoid arthritis were randomized to one of the two treatment groups and followed over time.

The outcome variable is a global impression scale (*Arthritis Categorical Scale*) measured at baseline (month 0), month 2, month 4, and month 6. This is a self-assessment of a patient's current arthritis, measured on a 5-level ordinal scale: (1) very good, (2) good, (3) fair, (4) poor, and (5) very poor. Baseline data on this outcome variable are available for 303 of the patients who participated in this trial; follow-up data at 6 months are available for 294 patients.

Variable List:
ID, Treatment (0=placebo, 1=auranofin therapy), baseline age (years), Arthritis Categorical Scale month 0, Arthritis Categorical Scale month 2, Arthritis Categorical Scale month 4, Arthritis Categorical Scale month 6.

## Objective
Goal: assess changes in the odds of a more favorable response over the duration of the study, and determine whether treatment with auranofin has an influence on these changes.

Letting $Y_{ij}$ denote the ordinal response for the $i^{th}$ subject at the $j^{th}$ occasion, we assume that the log odds of a more favorable response at each occasion follows the proportional odds model
$$
log\{\frac{Pr(Y_{ij} \leq k)}{Pr(Y_{ij}>k)}\}=\alpha_k + \beta_1Trt_i + \beta_2\sqrt{Month_{ij}} + \beta_3Trt_i \times \sqrt{Month_{ij}}
$$


```{r}
#Load data:
#library(foreign)
arthritis <- read.dta("arthritis.dta")

#transform data into long format
arthritis_long <- arthritis %>% pivot_longer(
  cols=starts_with("y"),
  names_to = "time",
  names_prefix = "y",
  values_to = "y",
  values_drop_na = TRUE) #drop missing values

#attach(arthritis_long)
arthritis_long$time <- as.numeric(arthritis_long$time)
arthritis_long$y <- as.factor(arthritis_long$y)
arthritis_long$month <- 2*(arthritis_long$time-1)
arthritis_long$sqrtmonth <- arthritis_long$month^0.5
```

## Ordinal responses
```{r}
#library(geepack)
summary(ordgee(ordered(y) ~ trt + sqrtmonth + trt:sqrtmonth, id = id, 
                waves=time, corstr=("independence"), data = arthritis_long))

summary(ordgee(ordered(y) ~ trt + sqrtmonth + trt:sqrtmonth, id = id, 
               waves=time, corstr=("unstructured"), data = arthritis_long))
```

## Dichotomous response variable
* If the Arthritis Categorical Scale is (1) very good or (2) good, the new response variable y1 = 1. 
* If the Arthritis Categorical Scale is (3) fair or (4) poor or (5) very poor, the new response variable y1 = 0.

```{r}
arthritis_long <- arthritis_long %>% 
  mutate(y1 = case_when(y==1 | y==2 ~ 1, #very good | good
                        y==3 | y==4 | y==5 ~ 0)) #fair | poor | very poor
arthritis_long$y1 <- as.factor(arthritis_long$y1)



model1 <- ordgee(ordered(y1) ~ trt + sqrtmonth + trt:sqrtmonth, 
                 data = arthritis_long, id = id, 
                 waves = time, mean.link = "logit", 
                 corstr = "independence")
summary(model1)
 
#SAME AS:
gee.in <- geeglm(y1 ~ trt + sqrtmonth + trt:sqrtmonth, data = arthritis_long,
              family = "binomial", id=id, 
              corstr = "independence", scale.fix = TRUE)
summary(gee.in)
QIC(gee.in)
```



